---
title: 'Interaction Effects from Regression Models: Plotting & Simple Slopes'
output: 
  html_document:
    toc: true
    toc_float: false
    includes:
      after_body: footer2.html
---


<font size="3">

```{r echo = F}
knitr::opts_chunk$set(message = F, cache = T, warning = F)  
```

<br>

This demo walks through how to plot interaction effects from regression models in R, complete with raw data points and 95% CI, and how to perform tests of simple slopes. 

<br>



```{r libraries, echo = TRUE, results = 'hide', message = FALSE}
library(car)
library(ggplot2)
```

<br>

### Load example dataset (from `car` package)
For this demo, we will use the Moore dataset from the car package. For details on this dataset, run `?Moore`. As the documentation from the package indicates, "the data are for subjects in a social-psychological experiment, who were faced with manipulated disagreement from a partner of either of low or high status. The subjects could either conform to the partner's judgment or stick with their own judgment."


```{r load data}
d <- Moore
```

<br>


### Variables and Research Question
Here, our `y` variable is `conformity`. Our `x` variable is `partner.status` (low vs. high), and our `z` variable (moderator) is authoritarianism (`fscore`). 

Our research question is whether the effect of `partner.status` on `conformity` depends on `fscore` (authoritarianism). We will, for the purposes of this demo, not concern ourselves with issues of power or sample size since our focus is on plotting and extracting simple effects. 

<br>

### Fit Regression Model

Before fitting the model, we will effect code the `partner.status` variable. It is not entirely necessary to do this, but it will simplify our interpretation of the results and facilitate plotting and analysis of the simple slopes in future steps. 

```{r}
d$partner.eff <- ifelse(d$partner.status == "low", -.5, .5)
```


We will also mean center `fscore`. We will keep `fscore` in its raw units (i.e., we will not standardize this variable). 

```{r}
d$fscore.c <- as.vector(scale(d$fscore, center = T, scale = F))
```

Now, we will fit a regression model in which we will include main effect terms for `fscore.c` and `partner.eff`, as well an interaction term. 

```{r fit regression model}
fit <- lm(conformity ~ fscore.c * partner.eff, data = d)
summary(fit)
```

We see that there is no main effect of `fscore.c`. In other words, averaging across conditions, authoritarianism is unrelated to conformity. We see that there is a main effect of `partner.eff`, indicating that people show more conformity to high (vs. low) status partners, controlling for authoritarianism. Importantly, however, we see that there is an interaction of these variables. Arguably the best way to interpret this interaction effect is to plot it and to perform simple slope analyses (i.e., compute the slope of `fscore.c` in each condition).


<br>

### Generate Model Predictions

We will now create a publication-quality visualization of our interaction effect. 

Note that there are several helper functions available in R that can help you visualize interaction effects faster. For example, the `effects` package can accomplish this in one line of code, as shown below. However, we are still missing elements we might want to include, such as the raw data points and confidence bands. 

```{r effectpackage example}
library(effects)
plot(effect(term = "fscore.c:partner.eff",
            mod = fit,
            default.levels=2),
     multiline=TRUE)
```

<br>

#### Create Dataframes for Each Level of IV

Next, we will generate dataframes for each level of the IV that we want to plot (two in this case, one of each level of partner status). Here, we will plot a separate line for each level of partner status and show levels of authoritarianism along the x-axis. One could swap the presentation of these variables. However, the present approach seemed like a good place to start because one variable is continuuous and the other is binary. From a plotting and analysis perspective, your X and moderator variables are interchangeable (although they might not be interchangeable theoretically).

First, let's create a dataframe that contains authoritarianism scores and fixes partner status to be high (`partner.eff == .5`). We will use this dataframe to feed values into the `predict()` function to get the model-predicted values for our y variable, as well as lower and upper bound estimates. 
If you had covariates, you would have mean centered them in the model and then filled them in here with the mean (0) for all.

```{r new dataframe high}
dfhigh <- data.frame(
  fscore.c = seq(min(d$fscore.c), max(d$fscore.c), .1), 
  # populate the column of your X of interest with a sequential order of values; 
  # alternatively, you could just specify the min and max.
  # .1 is arbitrary, but just specifies that R should fill in values
  # between min and max in .1 increments. Helps CIs look smooth in plotting.
  
  partner.eff = .5  # let's set "high" value of moderator +.5
  )  
```

We repeat the same procedure, but fix the value of partner status to be low (`partner.eff == -.5`).

```{r new dataframe low}
dflow <- data.frame(
  fscore.c = seq(min(d$fscore.c), max(d$fscore.c), .1), 
  partner.eff = -.5  
  )  
```

<br>

#### Model Predictions

We will now use the `predict()` function to generate fitted values. We will append these values to the dataframes we created in the step above. 

```{r predicted values}
partnerhigh <- cbind(dfhigh, predict(fit, dfhigh, interval = "confidence", type =c("response", "terms")))

partnerlow <- cbind(dflow, predict(fit, dflow, interval = "confidence", type =c("response", "terms")))

```

<br>

### Plotting
Now, let's use these predicted values in a graph using the `ggplot2()` package. 

```{r}
intplot <- ggplot(partnerlow, aes(fscore.c, fit)) +
  geom_point(data = d, # use original data to plot datapoints 
             aes(fscore.c, conformity, shape = partner.status, color = partner.status), 
             alpha = .7, size = 2, show.legend = T) +
  
  scale_color_manual(values=c("high" = "blue", 
                              "low" = "red"), 
                     name = "Partner Status") +
  
  scale_shape_manual(values=c("high" = 19, 
                              "low" = 1), 
                     name = "Partner Status") +
  labs(x = "Authoritarianism (fscore.c)", y = "Conformity", title = " ")  + 
  
  geom_ribbon(data = partnerlow, aes(ymin=lwr, ymax=upr), alpha = .3, fill = "red") +
  geom_line(data = partnerlow, aes(fscore.c, fit), size = 1.5, color = "red", linetype = "dashed") +
  
  geom_ribbon(data = partnerhigh, aes(ymin=lwr, ymax=upr), alpha = .3, fill = "blue") +
  geom_line(data = partnerhigh, aes(fscore.c, fit), size = 1.5, color = "blue") +
  theme_bw() 
intplot
```

This plot suggests that there is a positive effect of authoritarianism for participants in the low partner status condition, and a negative effect for those in the high status condition. Our next step is to determine whether these simple slopes differ significantly from 0. 

<br>

### Simple Slope Analysis

To perform tests on the simple slopes (the slopes of authoritarianism in each condition), we will use a function that I wrote. 

```{r simpleslope function}
condslope <- function(x, z, c, y){
  # x = name of x variable
  # z = name of z variable (moderator)
  # c = conditional value of z
  # y = reg object
  # reg function must be in this order x + z + x*z
  out <- summary(y)
  xz <- paste(x, z, sep=":")
  w0.intercept <- out$coefficients["(Intercept)", "Estimate"] + ((out$coefficients[z, "Estimate"])*c)
  w1.slope <- out$coefficients[x,"Estimate"] + ((out$coefficients[xz,"Estimate"])*c)
  #y.cond <- w0.intercept + w1.slope*xvalue
  
  coef2.var <- out$coef[x,"Std. Error"]^2
  coef4.var <- out$coef[xz,"Std. Error"]^2
  out.vcov <- vcov(y)
  cond.se <- sqrt(coef2.var + (c) * (c) * coef4.var + 2 * (c) * out.vcov[x, xz])
  
  t.val <- w1.slope/cond.se
  p.val <- 2*(1-pt(abs(t.val), y$df, lower.tail=T))
  lower95 <- w1.slope-qt(0.975,y$df)*cond.se
  upper95 <- w1.slope+qt(0.975,y$df)*cond.se
  
  return(list(w0.intercept=round(w0.intercept, digits = 2), 
              w1.slope=round(w1.slope, digits = 2), 
              t.val=round(t.val, digits = 2), 
              p.val=round(p.val, digits = 3), 
              lower95=round(lower95, digits = 2), 
              upper95=round(upper95, digits = 2)))
}

```

First, let's test the slope of authoritarianism in the high partner status condition.

```{r}
condslope("fscore.c", "partner.eff", .5, fit)
```

We see that the slope of authoritarianism in this condition is *b* = `r condslope("fscore.c", "partner.eff", .5, fit)$w1.slope`. Given the *t*, *p*, and 95% CI, we can conclude that this slope is significantly different from 0. 

Now, we will run the equivalent test for the lower partner status condition. 

```{r}
condslope("fscore.c", "partner.eff", -.5, fit)
```

Although the slope is positive, *b* = `r condslope("fscore.c", "partner.eff", -.5, fit)$w1.slope`, we are unable to conclude that it differs significantly from 0. At most, it is "trending" or "marginal." 

</font> 
<font size="2"> 
*[View .Rmd source code](/Users/zeekatherine/Desktop/WorkSchool/kzee.github.io/Interactions_Demo.Rmd){target="_blank"}*
</font> 
<font size="2"> 
\
*updated March 19, 2019*
</font> 