---
title: 'Spaghetti Plot for Multilevel Logistic Regression'
output: 
  html_document:
    toc: true
    toc_float: false
    includes:
      after_body: footer2.html
---


<font size="3">

```{r echo = F}
knitr::opts_chunk$set(message = F, cache = T, warning = F)  # Code not shown by default
```

# Overview
This demo shows how to create a spaghetti plot of individual-specific effects from a multilevel logistic model. 

We will use Ch. 6 data from Bolger & Laurenceau's *Intensive Longitudinal Methods* book. In this chapter, the authors used the example of the potential effect of a female partner's morning anger (`amangcw`) on her male partner's report of conflict (`pconf`; yes/no) later that evening. We will fit a similar model to the model presented in the book, except we will use Bayesian estimation. Using a Bayesian model will allow us to model random slopes. 

<br>

# Read in Data
We will read in the data directly from the *Intensive Longitudinal Methods* [website](http://www.intensivelongitudinal.com/). 
```{r pull in dataset}
library(data.table)
c <- as.data.frame(
  fread("curl http://www.intensivelongitudinal.com/ch6/ch6R.zip | tar -xf- --to-stdout *categorical.csv")
)

head(c)

```
<br>

# Load Libraries
```{r load packages, results = "hide", message=F, warning = F}
library(brms)
library(ggplot2)
library(dplyr)
```

<br>

# Fit Model 
```{r fit model, results = "hide", message = F, warning=F}
catmod1B <- brm(pconf ~ amangcw + amangcb + lpconfc + time7c + 
                (amangcw | id), 
              family = bernoulli, 
              data = c, cores=4) 

```

```{r model summary}
summary(catmod1B)
```
<br>

# Generate Predictions
<br>

## Create a new dataframe

```{r}
df1 <- dplyr::select(c, id, amang, amangcw, pconf) 
df1$amangcb <- 0
df1$time7c <- 0
df1$lpconfc <- 0
```
<br>

## Use new dataframe to generate fitted values
```{r fitted for original data}

fitdf1 <- cbind(df1, fitted(catmod1B, newdata = df1, re_formula = NULL,  incl_autocor = F)) 

```
<br>

## Obtain the Fixed effect
```{r fixed eff}

x <- data.frame(amangc = seq(min(c$amangc), max(c$amangc), by = .1),
                amang = seq(min(c$amang), max(c$amang), by = .1))

x$predM <-(1/(1 + exp(-(fixef(catmod1B)["Intercept", "Estimate"] + 
                          fixef(catmod1B)["amangcw", "Estimate"]*x$amangc))))
```
<br>

# Plot the Results
<br>


## Plot Individual Specific Effects
```{r}
catspag0 <- ggplot(fitdf1, aes(amang, Estimate, group = id)) + 
  theme_bw() + 
  geom_line() + 
  ylab("Male Report of Conflict") + 
  xlab("Female Morning Anger") +
  scale_y_continuous(breaks = c(0, .5, 1))  
catspag0
```
<br>

## Add in Fixed (Average) Effect

```{r}
catspag0 + geom_line(data = x, aes(amang, predM, group = NULL), 
            color = "blue",
            size = 2)
```

<br>

So we have our plot. It's not bad, but it's not great either. The individual-specific lines look a bit spikey, and one even looks crooked. We are missing that smooth, sigmoidal shape that characterizes logistic regression effects. 

The reason for this is that we need to add more values for each person in order to properly display their effect. As of now, ggplot simply connects the points available, resulting in effects that appear too linear or kinked in some way. 

<br>

# Interpolate Data
To remedy the "spikey" way our spaghetti plot looks, we will need to interpolate data. To be clear, we are NOT adding more data to the model; the model only uses our actual observations to generate estimates. Instead, we are just filling in more datapoints for the purposes of plotting, and we will do this only within each subject's observed range. In other words, we will not add data that is fall outside of each person's mininum observed value or maximum observed value. 

To accomplish this, we will need to create a new dataframe, called `df2`. We will begin by setting this dataframe to `NULL` so that it is just a placeholder waiting to "catch" the data we will populate it with. 

Next, we will set up a for loop in which we will ask R to take each subject's minimum and maximum observed value, and generate values within that range in small increments, in this case increments of .05 (as shown with the `seq()` function). This will enable us to have a high degree of granularity so that our curves will be smooth when we plot them. We will also need to set other variables to 0, as they were mean centered in our model. 


```{r}
df2 <- NULL
for (i in unique(c$id)) {
  cseq <- data.frame(
  amangcw = seq(min(subset(c, id==i)$amangcw, na.rm=T), 
               max(subset(c, id==i)$amangcw, na.rm=T), .05), 
  amang = seq(min(subset(c, id==i)$amang, na.rm=T), 
               max(subset(c, id==i)$amang, na.rm=T), .05), 
  amangc = seq(min(subset(c, id==i)$amangc, na.rm=T), 
              max(subset(c, id==i)$amangc, na.rm=T), .05), 
  amangcb = 0, 
  lpconfc = 0, 
  time7c = 0, 
  id = i)
  df2 <- rbind(df2, cseq)
}

```
<br>

Now we have a very fine-grained version of our dataframe.

```{r}
head(df2)
```

<br>

Next, we will use the `fitted()` function to generate fitted values using this new, fine-grained dataframe. As noted in my Plotting Fixed Effects demo, the `brms` package that we used to fit our model features both `fitted()` and `predict()` to generate model predictions, but these function do slightly different things. We will use `fitted()` here; however, we would get the same thing using `predict()` in this case because we are not including uncertainty (i.e., credibility intervals) in these plots.

We will create a new dataframe, `cseqpred`, in which we will take our new dataframe, `interp`, and append to it the model predicted values. 

```{r fitted with new df, message = F, warning = F} 
fitdf2 <- cbind(df2, fitted(catmod1B, newdata = df2, re_formula=NULL, incl_autocor = F))
head(fitdf2)
```

<br>

Now, we can create a plot very similar to our prior plot, but we will instead use our more fine-grained dataset so we get that sigmoidal curve we are after. 

```{r}
catspag <- ggplot(fitdf2, aes(amang, Estimate, group = id)) + 
  ylab("Male Report of Conflict") + 
  xlab("Female Morning Anger") +
  theme_bw() +  
  scale_y_continuous(breaks = c(0, .5, 1)) +
  geom_line() + 
  geom_line(data = x, aes(amang, predM, group = NULL), 
            color = "blue",
            size = 2) 
catspag
```

<br>

That looks a lot better!

<br>

# Side by Side Comparison

As a final step, let's compare these two plots side by side so we can see how using the interpolated data has improved our plot.

```{r side by side plots, message = F, warning = F}
library(gridExtra)
grid.arrange(catspag0 + geom_line(data = x, aes(amang, predM, group = NULL), 
            color = "blue",
            size = 2) + labs(title = "Without Interpolated Data"), 
            
            catspag + labs(title = "With Interpolated Data"), ncol = 2)

```

</font> 

<br>
<font size="2"> 
*[View .Rmd source code](/Users/zeekatherine/Desktop/WorkSchool/kzee.github.io/LogisticSpaghetti_Demo.Rmd){target="_blank"}*
</font> 

<font size="2"> 
*updated April 22, 2019*
</font> 